Multilevel spin models in cold atomic systems

### 2021-04-01 sun_phases/spin_model_paper/main.tex
add back the *correct* 'back-to-back' / 'XX' state

If the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, then the mode-changing collisions in $H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work).
The operator content of terms with $k=\ell$ and $p=q$ takes the form $c_{\ell\mu}^\dag c_{\ell\mu} c_{q\nu}^\dag c_{q\nu}$, so these terms contribute an overall energy $\frac12NU$ that can be neglected in the absence of coherent atom number fluctuations.
Defining the spin operators $s_{\mu\nu q}\equiv c_{q\mu}^\dag c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians become
\begin{align}
  H_{\t{kin}} &= -J\sum_{q,\mu} \cos\p{q} s_{\mu\mu q},
  \label{H_kin_start} \\
  H_{\t{int}} &= -\f{U}{2N} \sum_{p,q,\mu,\nu} s_{\mu\nu p} s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case, we can neglect the fermionic origins of $H_{\t{int}}$ and instead treat our system as a bona fide spin model.
Note that the ``kinetic'' terms of this spin model ($H_{\t{kin}}$) are proportional to the identity operator, and can be neglected for the time being.
The fermionic origin of these terms will return to play an important role when we consider spin-orbit coupling in Section \ref{sec:SOC}.

@@

We now discuss the spin Hamiltonian $H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v s_p\c\v s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v s_p\c\v s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v s_p\c\v s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns (distinct) definite energies to anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.

By summing over all pair-wise exchange terms $\v s_p\c\v s_q$, the interaction Hamiltonian $H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $U$ between the manifold of all permutationally-symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
In the case of SU(2), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice can tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U_0$.
(b,c) When the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, the frozen-mode approximation enables writing the interaction Hamiltonian as a spin model consisting of exchange terms $\v s_p\c\v s_q$, which swap the states two spins pinned to momentum modes $p,q$.
(d) Interactions thereby open an energy gap $U$ between the manifold of permutationally-symmetric states, and the orthogonal complement of states that break permutational symmetry.
}
\label{fig:spin_model}
\end{figure}

@@

In principle, making the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ also transforms the interaction Hamiltonian as
\begin{align}
  H_{\t{int}} \to H_{\t{int}}^{(\phi)}
  \equiv -\f{U}{2N} \sum_{p,q,\mu,\nu}
  c_{p-\mu\phi,\mu}^\dag c_{p-\nu\phi,\nu}
  c_{q-\nu\phi,\nu}^\dag c_{q-\mu\phi,\mu}.
\end{align}
However, $H_{\t{int}}^{(\phi)}$ cannot be exactly reduced to a spin Hamiltonian that addresses the same spin degrees of freedom as $H_{\t{kin}}^{(\phi)}$.
Nonetheless, for sufficiently weak SOC (small $s\phi$) the effect of $H_{\t{int}}^{(\phi)}$ may be approximated by $H_{\t{int}}^{(0)}=H_{\t{int}}$, which has previously been benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}.
To ensure that $H_{\t{kin}}^{(\phi)}$ does not become trivial as $\phi\to0$, we can simultaneously take $J\to\infty$ ($J$ large compared to $U$) while keeping $J\phi$ constant.
Altogether, the interacting spin Hamiltonian in the gauged frame becomes
\begin{align}
  H_{\t{spin}} = -\f{U}{2N} \v S\c\v S + J\phi \sum_q \sin\p{q} s_{\z,q}.
  \label{eq:H_spin}
\end{align}

@@

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ prior to evolution under the ``gauged frame'' spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin} or the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
We assume the capability to prepare an initial state in which all spins are maximally polarized along $\v v_+$, i.e.~$\ket{\v v_+}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauged frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, by choosing $H_{\t{drive}}^{\t{single}}\propto s_\x^2$ (see Table \ref{tab:drives}), we can transform the polarized state $\ket{s}^{\otimes N}$ into $\p{\ket{s}+\ket{-s}}^{\otimes N}$, where every spin is in a coherent superposition of states pointing in opposite directions along $\v v_+$.
This state has a mean spin vector that rapidly vanishes with a growing spin dimension $n=2s+1$, $\bk{S_\y}=\bk{S_\z}=0$ and $\bk{S_\x}=N\times 2s/2^{2s}=O(s/4^s)$ (where $S_\alpha = \sum_j s_{\alpha j}$), but a growing axial spin variance $\bk{S_\z^2}=Ns^2$.

@@

\begin{figure}
\centering
\includegraphics{mean_op_ss_X.pdf}
\caption{
Long-time averages of $\bar\sigma_\x$ and $\ss$ for different spin dimensions $n$ (indicated in the legend) as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket\x^{\otimes N}$ up to a time $T=10^5/U$.
Insets show same data after normalizing $J\phi/U\to J\phi/U \times (n/2)^{1/3}$, as well as transforming $\bbk{\bar\sigma_\x}_\MF$ and $\bbk{\ss}_\MF$ according to Eq.~\eqref{eq:rescale}.
}
\label{fig:mean_op_ss_X}
\end{figure}

Figure \ref{fig:mean_op_ss_X} shows the long-time averages of the transverse magnetization $\bar\sigma_\x$ and interaction energy $\ss$ as computed by mean-field simulations of $N=100$ spins in an initial $x$-polarized state $\ket\x^{\otimes N}$, where
\begin{align}
  \ket\x \equiv e^{-\i\frac{\pi}{2}s_\y}\ket{s}
  = \f1{2^s} \sum_\mu { 2s \choose s+\mu }^{1/2} \ket{\mu}.
  \label{eq:state_x}
\end{align}
Here ${m \choose k}$ is a binomial coefficient.
As expected, the spin model exhibits a mean-field dynamical phase transition between an ordered phase at small $J\phi/U$ and a disordered phase at large $J\phi/U$.
The ordered phase has a non-zero transverse magnetization $\bbk{\sigma_\x}_\MF$ and a large interaction energy $\bbk{\ss}_\MF$, while the disordered phase has no transverse magnetization and an interaction energy close to its minimal value.
Numerically, we find that the transition between these phases occurs at a critical field $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ with $\alpha\approx1/3$ (see Figure \ref{fig:crit_fields_X}).

>>>>

If the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, then the mode-changing collisions in $H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work)\footnote{Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $c_{\pi-p,\mu}^\dag c_{\pi-q,\mu} c_{p\nu}^\dag c_{q\nu}$, which conserve both momentum and energy.
We defer a careful treatment of these terms to future work, noting only that they vanish on the manifold of permutationally-symmetric spin states with one atom per lattice site.}.
The operator content of terms with $k=\ell$ and $p=q$ takes the form $c_{\ell\mu}^\dag c_{\ell\mu} c_{q\nu}^\dag c_{q\nu}$, so these terms contribute an overall energy $\frac12NU$ that can be neglected in the absence of coherent atom number fluctuations.
Defining the spin operators $s_{\mu\nu q}\equiv c_{q\mu}^\dag c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians become
\begin{align}
  H_{\t{kin}} &= -J\sum_{q,\mu} \cos\p{q} s_{\mu\mu q},
  \label{H_kin_start} \\
  H_{\t{int}} &= -\f{U}{2N} \sum_{p,q,\mu,\nu} s_{\mu\nu p} s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case, we can neglect the fermionic origins of $H_{\t{int}}$ and instead treat our system as a bona fide spin model.
Note that the ``kinetic'' terms of this spin model ($H_{\t{kin}}$) are proportional to the identity operator, and can be neglected for the time being.
The fermionic origin of these terms will return to play an important role when we consider spin-orbit coupling in Section \ref{sec:SOC}.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites can tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U_0$.
(b,c) When the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, the frozen-mode approximation enables writing the interaction Hamiltonian as a spin model consisting of exchange terms $\v s_p\c\v s_q$, which swap the states two spins pinned to momentum modes $p,q$.
(d) Interactions thereby open an energy gap $U=U_0\times N/L$ between the manifold of permutationally-symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v s_p\c\v s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v s_p\c\v s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v s_p\c\v s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns (distinct) definite energies to anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\v s_p\c\v s_q$, the interaction Hamiltonian $H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $U$ between the manifold of all permutationally-symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

In the case of SU(2), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

@@

@@

In principle, making the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ also transforms the interaction Hamiltonian as $H_{\t{int}} \to H_{\t{int}}^{(\phi)}$, where
\begin{align}
  H_{\t{int}}^{(\phi)}
  \equiv -\f{U}{2N} \sum_{p,q,\mu,\nu}
  c_{p-\mu\phi,\mu}^\dag c_{p-\nu\phi,\nu}
  c_{q-\nu\phi,\nu}^\dag c_{q-\mu\phi,\mu}.
\end{align}
However, $H_{\t{int}}^{(\phi)}$ cannot be exactly reduced to a spin Hamiltonian that addresses the same spin degrees of freedom as $H_{\t{kin}}^{(\phi)}$.
Nonetheless, for sufficiently weak SOC (small $s\phi$) the effect of $H_{\t{int}}^{(\phi)}$ may be approximated by $H_{\t{int}}^{(0)}=H_{\t{int}}$, which has previously been benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}.
To ensure that $H_{\t{kin}}^{(\phi)}$ does not become trivial as $\phi\to0$, we can simultaneously take $J\to\infty$ ($J$ large compared to $U$) while keeping $J\phi$ constant.
Altogether, the interacting spin Hamiltonian in the gauged frame becomes
\begin{align}
  H_{\t{spin}} = -\f{U}{2N} \v S\c\v S + J\phi \sum_q \sin\p{q} s_{\z,q}.
  \label{eq:H_spin}
\end{align}

@@

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ prior to evolution under the ``gauged frame'' spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin} or the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
We assume the capability to prepare an initial state in which all spins are maximally polarized along $\v v_+$, i.e.~$\ket{\v v_+}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauged frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, we can prepare an $N$-fold product of the state
\begin{align}
  \ket{\x\x} \equiv e^{-\i\frac{\pi}{2}s_\y^2} \ket{s}
  \simeq \ket{+\x} + \ket{-\x},
\end{align}
where $\simeq$ denotes equality up to a global phase and normalization factor, and
\begin{align}
  \ket{\pm\x} \equiv e^{\mp\i\frac{\pi}{2}s_\y} \ket{s}
  = \f1{2^s} \sum_\mu \p{\pm1}^{s-\mu}
  { 2s \choose s+\mu }^{1/2} \ket{\mu}
\end{align}
are states of a single spin polarized in opposite directions along the $x$ axis.
Here ${m \choose k}$ is a binomial coefficient.
The state $\ket{\x\x}$ has a vanishing mean spin vector, $\bk{S_\x}=\bk{S_\y}=\bk{S_\z}=0$, (where $S_\alpha = \sum_j s_{\alpha j}$), but variances $\bk{S_\x^2}=N\times s^2$ and $\bk{S_\y^2}=\bk{S_\z^2}=N\times s/2$.

@@

\begin{figure}
\centering
\includegraphics{mean_op_ss_X.pdf}
\caption{
Long-time averages of $\bar\sigma_\x$ and $\ss$ for different spin dimensions $n$ (indicated in the legend) as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket{+\x}^{\otimes N}$ up to a time $T=10^5/U$.
Insets show same data after normalizing $J\phi/U\to J\phi/U \times (n/2)^{1/3}$, as well as transforming $\bbk{\bar\sigma_\x}_\MF$ and $\bbk{\ss}_\MF$ according to Eq.~\eqref{eq:rescale}.
}
\label{fig:mean_op_ss_X}
\end{figure}

Figure \ref{fig:mean_op_ss_X} shows the long-time averages of the transverse magnetization $\bar\sigma_\x$ and interaction energy $\ss$ as computed by mean-field simulations of $N=100$ spins in an initial $x$-polarized state $\ket{+\x}^{\otimes N}$.
As expected, the spin model exhibits a mean-field dynamical phase transition between an ordered phase at small $J\phi/U$ and a disordered phase at large $J\phi/U$.
The ordered phase has a non-zero transverse magnetization $\bbk{\sigma_\x}_\MF$ and a large interaction energy $\bbk{\ss}_\MF$, while the disordered phase has no transverse magnetization and an interaction energy close to its minimal value.
Numerically, we find that the transition between these phases occurs at a critical field $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ with $\alpha\approx1/3$ (see Figure \ref{fig:crit_fields_X}).
